### 2016-10-04 Chapters/SolvationAndCorrelation.tex
Adjust equations

% mainfile: ../RobertoDiRemigioPhDThesis.tex
%************************************************
\chapter{Electron Correlation and Solvation}\label{ch:solvation-correlation}

\begin{epigraphs}
\qitem{Potius sero quam numquam}{--- \textsc{Titus Livius}, \textit{Ab Urbe Condita}}
\end{epigraphs}

I will present results on the formulation of quantum/classical
polarizable models when the quantum part of the system includes a
many-body description of electron correlation either by a \gls{MBPT},
\gls{CC} or hybrid approach.
The contents of this Chapter are the result of an ongoing collaboration
with prof. T. Daniel Crawford (Virginia Tech) and Dr. Andrew
C. Simmonett (NIH)

@@

Sections \ref{sec:effective-cc-lagrangian} and \ref{sec:source-terms}
will introduce the effective \acrshort{CC} Lagrangian and the relevant
notation for the source terms in the classical energy functional,
respectively.
In Section \ref{sec:pcm-cc-models} I will derive the governing equation
for the effective Lagrangian and show how previous work on
continuum\autocite{Christiansen1999-tj, Cammi2009-gu, Caricato2011-tx}
and explicit models\autocite{Kongsted2003-py, Sneskov2011-jm} can easily
be derived in the variational framework.

@@

For ease of reference, I will repeat some of the notational conventions
introduced in Chapter \ref{ch:QM} here.
The cluster operator at truncation level $\mathcal{M}$ is:
\begin{equation}
  T = \sum_{u=1}^\mathcal{M} T_u =
  \sum_{u=1}^\mathcal{M}\sum_{\mu_u}\tamp{\mu_u}\cluster{\mu_u},
  \tag{\ref{eq:cluster} from Chapter \ref{ch:QM}}
\end{equation}
where $\tamp{\mu_u}$ and $\cluster{\mu_u}$ are the cluster amplitudes
and cluster operators for the $\mu_u$-th excited determinant at the
$u$-th excitation level, respectively.
The similarity transformed counterpart of the operator $O$ is denoted
by $\overline{O}$:
\begin{equation}
  \overline{O} =\expo{-T}O\expo{T}.
  \tag{\ref{eq:similarity-transformation} from Chapter \ref{ch:QM}}
\end{equation}
We recall that hermiticity of $O$ is not preserved by the similarity
transformation.
Manipulation of similarity transformed operators involves the use of
their \acrlong{BCH} expansions:
\begin{equation}
  \overline{O}
  =
  O + \BCHfirst{O}{T} + \BCHsecond{O}{T} + \ldots
  =
  O + \tilde{O},
  \tag{\ref{eq:bch} from Chapter \ref{ch:QM}}
\end{equation}
since we are concerned with one- and two-electron operators only,
such expansions will truncate after the third and fifth terms,
respectively.\autocite{Helgaker2000-tz}
The similarity transformation by means of the singles cluster operator
only, \ie the $T_1$-transformation, will be denoted by a "check" on top
of the operator:
\begin{equation}
  \check{O} = \expo{-T_1}O\expo{T_1}.
  \tag{\ref{eq:t1-transformation} from Chapter \ref{ch:QM}}
\end{equation}
The $T_1$-transformation preserves the particle rank of the operator
$O$, see Appendix \ref{app:mathematical-results} for details.
Finally, the $u$-th excited component of the \acrlong{CC} left state
will be denoted as:
\begin{equation}
  \bra{\tbar{u}} = \sum_{\mu_u}\tbar{\mu_u}\bra{\mu_u},
  \tag{\ref{eq:left-state} from Chapter \ref{ch:QM}}
\end{equation}
where $\tbar{\mu_u}$ are Lagrange multipliers.

@@

Once again, let us assume that the reference state is a closed-shell
\acrshort{HF} determinant. Further, we assume that the reference determinant
was optimized in the presence of the classical polarizable environment.
Then the polarization degree of freedom corresponding to the environment
satisfies:
\begin{equation}
  \V\p_\mathrm{HF} + \s_\mathrm{HF} = 0
\end{equation}
and corresponds to the \emph{reference reaction field}.
In the subsequent correlated treatment one can separate the reaction
field into reference and correlated components:
\begin{equation}
  \p^\mathrm{tot} = \p_\mathrm{HF} + \p
\end{equation}
and similarly for the source term:
\begin{equation}
  \begin{aligned}
    \s(\tamp{}, \tbar{})_\mathcal{M} &= \s_\mathrm{HF} + \s_{\mathrm{N}}(\tamp{}, \tbar{})_\mathcal{M} \\
                     &=
    \braket{\mathrm{HF} | \s | \mathrm{HF}}
  + \braket{\mathrm{HF} | \tilde{\s} | \mathrm{HF}}
  + \sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} | \overline{\s} | \mathrm{HF}}
  \end{aligned}
\end{equation}
The \acrshort{BCH} expansions of the source term has been rewritten as:
\begin{equation}
  \overline{\s} = \s + \tilde{\s} = \s + \commutator{\s}{T}
  + \frac{1}{2}\commutator{\commutator{\s}{T}}{T}
\end{equation}
taking advantage of the fact that $\s$ is a
nondiagonal, one-electron operators and hence its commutator
expansion truncates at the third term. Moreover, normal-ordering of the
operators has been introduced.\autocite{Crawford2000-ey, Shavitt2009-mr}
This is equivalent to applying a shift to the correlation part of the
source term one-electron operators to remove the reference source terms:
\begin{equation}
  \begin{aligned}
    \s_{\mathrm{N}}(\tamp{}, \tbar{})_\mathcal{M}
  &=
   \braket{\mathrm{HF} | \tilde{\s} | \mathrm{HF}}
  + \sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} | \overline{\s} | \mathrm{HF}} \\
  &=
    \braket{\mathrm{HF} | \overline{\s} - \s_\mathrm{HF} | \mathrm{HF}}
  + \sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} | \overline{\s} - \s_\mathrm{HF} | \mathrm{HF}}
  \end{aligned}
\end{equation}

Eventually, the polarization functional can be rewritten as:
\begin{equation}
  \begin{aligned}
  U_\mathrm{pol}
  &=
  \frac{1}{2}({}^t\p_\mathrm{HF}+{}^t\p)\V(\p_\mathrm{HF}+\p)
  + ({}^t\p_\mathrm{HF}+{}^t\p)(\s_\mathrm{HF}+\s_{\mathrm{N}}(\tamp{},\tbar{})_\mathcal{M}) \\
  &=
    \frac{1}{2}{}^t\p_\mathrm{HF}\V\p_\mathrm{HF} + {}^t\p_\mathrm{HF}\s_\mathrm{HF}
  + \frac{1}{2}{}^t\p_\mathrm{HF}\V\p + \frac{1}{2}{}^t\p\V\p_\mathrm{HF} \\
  &+ {}^t\p_\mathrm{HF}\s_{\mathrm{N}}(\tamp{},\tbar{})_\mathcal{M}
  + {}^t\p\s_\mathrm{HF}
  + \frac{1}{2}{}^t\p\V\p + {}^t\p\s_{\mathrm{N}}(\tamp{},\tbar{})_\mathcal{M} \\
  &=
    U_\mathrm{pol}^\mathrm{ref}
  + \frac{1}{2}{}^t\p\V\p + {}^t\p\s_{\mathrm{N}}(\tamp{},\tbar{})_\mathcal{M}
  + {}^t\p_\mathrm{HF}\s_{\mathrm{N}}(\tamp{},\tbar{})_\mathcal{M}
  \end{aligned}
\end{equation}
The reference polarization energy $U_\mathrm{pol}^\mathrm{ref}$ still
appears among the terms in the functional but will obviously not enter
in the optimization of the \acrshort{CC} wave function.
One can then write the effective Lagrangian as:
\begin{equation}
  \begin{aligned}
    \lag(\tamp{}, \tbar{}, \p)_\mathcal{M} &=
  \braket{\mathrm{HF} | \overline{H_0} | \mathrm{HF}}
  + \sum_{u=1}^{\mathcal{M}}\braket{\tamp{u} | \overline{H_0} | \mathrm{HF}} \\
  &+
  \frac{1}{2}{}^t\p\V\p + {}^t\p\s_{\mathrm{N}}(\tamp{},\tbar{})_\mathcal{M}
  + {}^t\p_\mathrm{HF}\s_{\mathrm{N}}(\tamp{},\tbar{})_\mathcal{M}
  + U_\mathrm{pol}^\mathrm{ref}
  \end{aligned}
\end{equation}
In accordance with previous work,\autocite{Cammi2009-gu,
Caricato2011-tx} the Hamiltonian $H_0$ is augmented with the
${}^t\p_\mathrm{HF}\s_{\mathrm{N}}(\tamp{},\tbar{})$ term to yield the
\acrlong{PTE} (\acrshort{PTE}) Hamiltonian -- $H$ -- and eventually the
\acrlong{PTED} (\acrshort{PTED}) Lagrangian:\autocite{Olivares_del_Valle1991-of,
Aguilar1991-vq, Olivares_del_Valle1991-tq, Olivares_del_Valle1993-xq,
Olivares_del_Valle1993-ra, Lipparini2009-io}
\begin{equation}\label{eq:pted-cc}
  \lag(\tamp{}, \tbar{}, \p)_\mathcal{M} =
  {}^\mathrm{PTE}\lag(\tamp{}, \tbar{})_\mathcal{M}
  + \frac{1}{2}{}^t\p\V\p + {}^t\p\s_{\mathrm{N}}(\tamp{},\tbar{})_\mathcal{M}
\end{equation}
where the \acrshort{PTE} Lagrangian has been introduced:
\begin{equation}
  {}^\mathrm{PTE}\lag(\tamp{}, \tbar{})_\mathcal{M}
  =
  \braket{\mathrm{HF} | \overline{H} | \mathrm{HF}}
  + \sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} | \overline{H} | \mathrm{HF}}
  + U_\mathrm{pol}^\mathrm{ref}
\end{equation}

As noted by \citeauthor{Cammi2009-gu} for the implicit \acrshort{PCM} model,
the \acrshort{PTE} Hamiltonian is computed in \acrshort{MP}-partitioned form by
employing the ``solvated'' orbitals and Fock matrix:
\begin{equation}
  H = F + \Phi.
\end{equation}
$F$ is a diagonal one-electron operator and $\Phi$ is the
fluctuation potential.
The simple nature of the Fock operator lets us re-express the \acrshort{PTE}
Lagrangian as follows:
\begin{equation}
  {}^\mathrm{PTE}\lag(\tamp{}, \tbar{})_\mathcal{M}
  =
  G_0
  + \sum_{u=1}^{\mathcal{M}}\tbar{\mu_u}\denom{\mu_u}\tamp{\mu_u}
  + \braket{\mathrm{HF} | \overline{\Phi} | \mathrm{HF}}
  + \sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} | \overline{\Phi} | \mathrm{HF}}
\end{equation}
where $G_0 = E_0 + U_\mathrm{pol}^\mathrm{ref}$ is the reference free
energy.

Before delving into the details of the governing equations for
amplitudes, multipliers and polarization, we need to take a closer look
at the source term in classical energy functional. This is the purpose
of the next Section.

@@

In the supermatrix formalism, the similarity-transformed terms will be
written as:
\begin{equation}
  \overline{\s} = \s + \tilde{\s} = \s +
  \commutator{\s}{T} +
  \frac{1}{2}\commutator{\commutator{\s}{T}}{T}
\end{equation}
It is important to single out which terms contribute in the various
expectation values where the similarity-transformed sources will be
involved.
This analysis is carried out in detail in Appendix \ref{sec:CC-expval}
and we report here the final results for the \gls{CCS}, \gls{CCSD} and
\gls{CCSDT} models.

For the \gls{CCS} model ($\mathcal{M} = 1$) one has:
\begin{equation}\label{eq:source-ccs}
  \begin{aligned}
    \s_{\mathrm{N}}(\tamp{}, \tbar{})_\mathrm{CCS} &=
  \braket{\mathrm{HF} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
  + \braket{\tbar{1} | \check{\s} | \mathrm{HF}} \\
  &=
  \braket{\mathrm{HF} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
  + \braket{\tbar{1} | \s | \mathrm{HF}} \\
  &+ \braket{\tbar{1} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
  + \frac{1}{2}\braket{\tbar{1} | \commutator{\commutator{\s}{T_1}}{T_1} | \mathrm{HF}}
  \end{aligned}
\end{equation}
while for the \gls{CCSD} model ($\mathcal{M} = 2$):
\begin{equation}\label{eq:source-ccsd}
  \begin{aligned}
  \s_{\mathrm{N}}(\tamp{}, \tbar{})_\mathrm{CCSD} &=
  \s_{\mathrm{N}}(\tamp{}, \tbar{})_\mathrm{CCS}
  + \braket{\tbar{1} | \BCHfirst{\s}{T_2} | \mathrm{HF}} \\
  &+ \braket{\tbar{2} | \BCHfirst{\s}{T_2} | \mathrm{HF}}
  + \braket{\tbar{2} | \commutator{\commutator{\s}{T_1}}{T_2} | \mathrm{HF}}
  \end{aligned}
\end{equation}
Eventually, within the \gls{CCSDT} model ($\mathcal{M} = 3$) four more
terms are added:
\begin{equation}\label{eq:source-ccsdt}
  \begin{aligned}
    \s_{\mathrm{N}}(\tamp{}, \tbar{})_\mathrm{CCSDT} &=
  \s_{\mathrm{N}}(\tamp{}, \tbar{})_\mathrm{CCSD}
  + \braket{\tbar{2} | \BCHfirst{\s}{T_3} | \mathrm{HF}}
  + \braket{\tbar{3} | \BCHfirst{\s}{T_3} | \mathrm{HF}} \\
  &+ \frac{1}{2}
  \braket{\tbar{3} | \commutator{\commutator{\s}{T_2}}{T_2} |
  \mathrm{HF}}
  + \braket{\tbar{3} | \commutator{\commutator{\s}{T_1}}{T_3} |
  \mathrm{HF}}
  \end{aligned}
\end{equation}

@@

Differentiation of the effective Lagrangian in Eq. \eqref{eq:pted-cc}
with respect to the variational parameters $\tamp{\mu_i}$,
$\tbar{\mu_i}$ and $\p$ yields the \acrshort{PTED}-\acrshort{CC}
equations:
\begin{subequations}\label{eq:pted}
  \begin{align}
   \tampEq{\mu_q}(\tamp{}, \tbar{}, \p)  &=
   \denom{\mu_q}\tamp{\mu_q} + \braket{\mu_q | \overline{\Phi} | \mathrm{HF}}
   + {}^t\p\braket{\mu_q | \overline{\s} | \mathrm{HF}}
             = 0 \label{eq:pted-cc-amplitudes}\\
   \tbarEq{\mu_q}(\tamp{}, \tbar{}, \p)
    &=
    \denom{\mu_q}\tbar{\mu_q} +
    \braket{\mathrm{HF} | \commutator{\overline{\Phi}}{\cluster{\mu_q}} | \mathrm{HF}} +
    \sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} |
    \commutator{\overline{\Phi}}{\cluster{\mu_q}} | \mathrm{HF}}
    \nonumber \\
    &+
    {}^t\p\braket{\mathrm{HF} | \commutator{\s}{\cluster{\mu_q}} | \mathrm{HF}}\delta_{\mu_q\mu_1} \nonumber \\
    &+
    {}^t\p\sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} | \commutator{\s}{\cluster{\mu_q}} | \mathrm{HF}}
    +
    {}^t\p\sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} | \commutator{\commutator{\s}{T}}{\cluster{\mu_q}} | \mathrm{HF}}
             = 0 \label{eq:pted-cc-multipliers}\\
    \Omega_\p(\tamp{}, \tbar{}, \p)
    &=
    \V\p + \s_{\mathrm{N}}(\tamp{}, \tbar{})_\mathcal{M} = 0
    \label{eq:pted-cc-polarization}
  \end{align}
\end{subequations}
A coupling of the amplitudes and multipliers equations is introduced by
the presence of the $\s_{\mathrm{N}}(\tamp{}, \tbar{})_\mathcal{M}$ expectation value
in the polarization equation.
This requires a proper macroiteration/microiteration self-consistency
scheme for its implementation.\autocite{Cammi2009-gu, Caricato2010-hx}
As a consequence, a single-point \acrshort{PTED}-\acrshort{CC} will suffer from a
$2\times$ prefactor in its computational cost with respect to an
\emph{in vacuo} \acrshort{CC} calculation.

It is thus convenient to devise approximations that are able to simplify the
Lagrangian and the governing equations by breaking or weakening the coupling.
While Eq. \eqref{eq:pted-cc-polarization} is \emph{directly} coupled to
the multipliers equation, since the source terms directly depends on
amplitudes and multipliers, it is only \emph{indirectly} coupled to the
amplitudes equation, where only the polarization enters.
Moreover, the leading terms in the correlated source operator
expectation value prominently involve the
singles cluster operator and the singles multiplier state.
Three approximations have currently been proposed: \gls{PTES}, \gls{PTE(S)}
and \gls{PTE}.

@@

  {}^{\mathrm{PTES}}\lag(\tamp{}, \tbar{}, \p)_\mathcal{M} =
  {}^\mathrm{PTE}\lag(\tamp{}, \tbar{})_\mathcal{M}
  + \frac{1}{2}{}^t\p\V\p + {}^t\p\s_{\mathrm{N}}(\tamp{},

>>>>

% mainfile: ../RobertoDiRemigioPhDThesis.tex
%************************************************
\chapter{Electron Correlation and Solvation}\label{ch:solvation-correlation}
\footnotetext{Potius sero quam numquam \\
\hspace*{15pt} --- \textsc{Titus Livius}, \\ \hspace*{15pt}
\textit{Ab Urbe Condita}}

I will present results on the formulation of quantum/classical
polarizable models when the quantum part of the system includes a
many-body description of electron correlation either by a \gls{MBPT},
\gls{CC} or hybrid approach.
The contents of this Chapter are the result of an ongoing collaboration
with prof. T. Daniel Crawford (Virginia Tech) and Dr. Andrew
C. Simmonett (NIH).

@@

Sections \ref{sec:effective-cc-lagrangian} and \ref{sec:source-terms}
will introduce the effective \acrshort{CC} Lagrangian and the relevant
notation for the source terms in the classical energy functional,
respectively.
In Section \ref{sec:pcm-cc-models} I will derive the governing equation
for the effective Lagrangian, showing how previous work on
continuum\autocite{Christiansen1999-tj, Cammi2009-gu, Caricato2011-tx}
and explicit models\autocite{Kongsted2003-py, Sneskov2011-jm} can easily
be derived in the variational framework.

@@

For ease of reference, I will repeat some of the notational conventions
introduced in Chapter \ref{ch:QM} here.
The cluster operator at truncation level $\mathcal{M}$ is:
\begin{equation}
  T = \sum_{u=1}^\mathcal{M} T_u =
  \sum_{u=1}^\mathcal{M}\sum_{\mu_u}\tamp{\mu_u}\cluster{\mu_u},
  \tag{\ref{eq:cluster} from Chapter \ref{ch:QM}}
\end{equation}
where $\tamp{\mu_u}$ and $\cluster{\mu_u}$ are the cluster amplitudes
and cluster operators for the $\mu_u$-th excited determinant at the
$u$-th excitation level, respectively.
The similarity transformed counterpart of the operator $O$ is denoted
by $\overline{O}$:
\begin{equation}
  \overline{O} =\expo{-T}O\expo{T}.
  \tag{\ref{eq:similarity-transformation} from Chapter \ref{ch:QM}}
\end{equation}
We recall that hermiticity of $O$ is not preserved by the similarity
transformation.
Manipulation of similarity transformed operators involves the use of
their \acrlong{BCH} expansions:
\begin{equation}
  \begin{aligned}
  \overline{O}
  &=
  O + \tilde{O} \\
  &=
  O + \BCHfirst{O}{T} + \BCHsecond{O}{T} + \ldots
  \end{aligned}
  \tag{\ref{eq:bch} from Chapter \ref{ch:QM}}
\end{equation}
since we are concerned with one- and two-electron operators only,
such expansions will truncate after the third and fifth terms,
respectively.\autocite{Helgaker2000-tz}
The similarity transformation by means of the singles cluster operator
only, \ie the $T_1$-transformation, will be denoted by a "check" on top
of the operator:
\begin{equation}
  \check{O} = \expo{-T_1}O\expo{T_1}.
  \tag{\ref{eq:t1-transformation} from Chapter \ref{ch:QM}}
\end{equation}
The $T_1$-transformation preserves the particle rank of the operator
$O$, see Appendix \ref{app:mathematical-results} for details.
Finally, the $u$-th excited component of the \acrlong{CC} left state
will be denoted as:
\begin{equation}
  \bra{\tbar{u}} = \sum_{\mu_u}\tbar{\mu_u}\bra{\mu_u},
  \tag{\ref{eq:left-state} from Chapter \ref{ch:QM}}
\end{equation}
where $\tbar{\mu_u}$ are Lagrange multipliers.

@@

Once again, let us assume that the reference state is a closed-shell
\acrshort{HF} determinant. Further, we assume that the reference determinant
was optimized in the presence of the classical polarizable environment.
Then the polarization degree of freedom corresponding to the environment
satisfies:
\begin{equation}
  \V\p_\mathrm{HF} + \s_\mathrm{HF} = 0
\end{equation}
and corresponds to the \emph{reference reaction field}.
In the subsequent correlated treatment one can separate the reaction
field into reference and correlated components:
\begin{equation}
  \p^\mathrm{tot} = \p_\mathrm{HF} + \p
\end{equation}
and similarly for the source term:
\begin{equation}
  \begin{aligned}
    \s(\tamp{}, \tbar{})_\mathcal{M} &= \s_\mathrm{HF} + \s_{\mathrm{N}}(\tamp{}, \tbar{})_\mathcal{M} \\
                     &=
    \braket{\mathrm{HF} | \s | \mathrm{HF}}
  + \braket{\mathrm{HF} | \tilde{\s} | \mathrm{HF}}
  + \sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} | \overline{\s} |
  \mathrm{HF}}.
  \end{aligned}
\end{equation}
The \acrshort{BCH} expansions of the source term has been rewritten as:
\begin{equation}\label{eq:source-BCH}
  \overline{\s} = \s + \tilde{\s} = \s + \commutator{\s}{T}
  + \frac{1}{2}\commutator{\commutator{\s}{T}}{T},
\end{equation}
taking advantage of the fact that $\s$ is a
nondiagonal, one-electron operators and hence its commutator
expansion truncates at the third term. Moreover, normal-ordering of the
operators has been introduced.\autocite{Crawford2000-ey, Shavitt2009-mr}
This is equivalent to applying a shift to the correlation part of the
source term one-electron operators to remove the reference source terms:
\begin{equation}
  \begin{aligned}
    \s_{\mathrm{N}}(\tamp{}, \tbar{})_\mathcal{M}
  &=
   \braket{\mathrm{HF} | \tilde{\s} | \mathrm{HF}}
  + \sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} | \overline{\s} | \mathrm{HF}} \\
  &=
    \braket{\mathrm{HF} | \overline{\s} - \s_\mathrm{HF} | \mathrm{HF}}
  + \sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} | \overline{\s} -
  \s_\mathrm{HF} | \mathrm{HF}}.
  \end{aligned}
\end{equation}

Eventually, the polarization functional can be rewritten as:
\begin{equation}
  \begin{aligned}
  U_\mathrm{pol}
  &=
  \frac{1}{2}({}^t\p_\mathrm{HF}+{}^t\p)\V(\p_\mathrm{HF}+\p) \\
  &+ ({}^t\p_\mathrm{HF}+{}^t\p)(\s_\mathrm{HF}+\s_{\mathrm{N}}(\tamp{},\tbar{})_\mathcal{M}) \\
  &=
    \frac{1}{2}{}^t\p_\mathrm{HF}\V\p_\mathrm{HF} + {}^t\p_\mathrm{HF}\s_\mathrm{HF}
  + \frac{1}{2}{}^t\p_\mathrm{HF}\V\p + \frac{1}{2}{}^t\p\V\p_\mathrm{HF} \\
  &+ {}^t\p_\mathrm{HF}\s_{\mathrm{N}}(\tamp{},\tbar{})_\mathcal{M}
  + {}^t\p\s_\mathrm{HF}
  + \frac{1}{2}{}^t\p\V\p + {}^t\p\s_{\mathrm{N}}(\tamp{},\tbar{})_\mathcal{M} \\
  &=
    U_\mathrm{pol}^\mathrm{ref}
  + \frac{1}{2}{}^t\p\V\p + {}^t\p\s_{\mathrm{N}}(\tamp{},\tbar{})_\mathcal{M}
  + {}^t\p_\mathrm{HF}\s_{\mathrm{N}}(\tamp{},\tbar{})_\mathcal{M}.
  \end{aligned}
\end{equation}
The reference polarization energy $U_\mathrm{pol}^\mathrm{ref}$ still
appears among the terms in the functional but will obviously not enter
in the optimization of the \acrshort{CC} wave function.
One can then write the effective Lagrangian as:
\begin{equation}
  \begin{aligned}
    \lag(\tamp{}, \tbar{}, \p)_\mathcal{M} &=
  \braket{\mathrm{HF} | \overline{H_0} | \mathrm{HF}}
  + \sum_{u=1}^{\mathcal{M}}\braket{\tamp{u} | \overline{H_0} | \mathrm{HF}} \\
  &+
  \frac{1}{2}{}^t\p\V\p +
  {}^t\p\s_{\mathrm{N}}(\tamp{},\tbar{})_\mathcal{M} \\
  &+ {}^t\p_\mathrm{HF}\s_{\mathrm{N}}(\tamp{},\tbar{})_\mathcal{M}
  + U_\mathrm{pol}^\mathrm{ref}
  \end{aligned}
\end{equation}
In accordance with previous work,\autocite{Cammi2009-gu,
Caricato2011-tx} the Hamiltonian $H_0$ is augmented with the
${}^t\p_\mathrm{HF}\s_{\mathrm{N}}(\tamp{},\tbar{})$ term to yield the
\acrlong{PTE} (\acrshort{PTE}) Hamiltonian -- $H$ -- and eventually the
\acrlong{PTED} (\acrshort{PTED}) Lagrangian:\autocite{Olivares_del_Valle1991-of,
Aguilar1991-vq, Olivares_del_Valle1991-tq, Olivares_del_Valle1993-xq,
Olivares_del_Valle1993-ra, Lipparini2009-io}
\begin{equation}\label{eq:pted-cc}
  \lag(\tamp{}, \tbar{}, \p)_\mathcal{M} =
  {}^\mathrm{PTE}\lag(\tamp{}, \tbar{})_\mathcal{M}
  + \frac{1}{2}{}^t\p\V\p +
  {}^t\p\s_{\mathrm{N}}(\tamp{},\tbar{})_\mathcal{M},
\end{equation}
where the \acrshort{PTE} Lagrangian has been introduced:
\begin{equation}
  {}^\mathrm{PTE}\lag(\tamp{}, \tbar{})_\mathcal{M}
  =
  \braket{\mathrm{HF} | \overline{H} | \mathrm{HF}}
  + \sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} | \overline{H} | \mathrm{HF}}
  + U_\mathrm{pol}^\mathrm{ref}.
\end{equation}

As noted by \citeauthor{Cammi2009-gu} for the implicit \acrshort{PCM} model,
the \acrshort{PTE} Hamiltonian is computed in \acrshort{MP}-partitioned form by
employing the \enquote{solvated} orbitals and Fock matrix:
\begin{equation}
  H = F + \Phi.
\end{equation}
$F$ is a diagonal one-electron operator and $\Phi$ is the
fluctuation potential.
The simple nature of the Fock operator lets us re-express the \acrshort{PTE}
Lagrangian as follows:
\begin{equation}
  \begin{aligned}
  {}^\mathrm{PTE}\lag(\tamp{}, \tbar{})_\mathcal{M}
  &=
  G_0
  + \sum_{u=1}^{\mathcal{M}}\tbar{\mu_u}\denom{\mu_u}\tamp{\mu_u}
  + \braket{\mathrm{HF} | \overline{\Phi} | \mathrm{HF}} \\
  &+ \sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} | \overline{\Phi} | \mathrm{HF}}
\end{aligned}
\end{equation}
where $G_0 = E_0 + U_\mathrm{pol}^\mathrm{ref}$ is the reference free
energy.

@@

Let us take a closer look at the source term in the classical energy
functional.
It is important to single out which terms contribute in the various
expectation values where the similarity-transformed sources Eq.
\eqref{eq:source-BCH} are involved.
This analysis is carried out in detail in Appendix \ref{sec:CC-expval}
and we report here the final results for the \gls{CCS}, \gls{CCSD} and
\gls{CCSDT} models.

For the \gls{CCS} model ($\mathcal{M} = 1$) one has:
\begin{equation}\label{eq:source-ccs}
  \begin{aligned}
    \s_{\mathrm{N}}(\tamp{}, \tbar{})_\mathrm{CCS} &=
  \braket{\mathrm{HF} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
  + \braket{\tbar{1} | \check{\s} | \mathrm{HF}} \\
  &=
  \braket{\mathrm{HF} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
  + \braket{\tbar{1} | \s | \mathrm{HF}} \\
  &+ \braket{\tbar{1} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
  + \frac{1}{2}\braket{\tbar{1} | \commutator{\commutator{\s}{T_1}}{T_1} | \mathrm{HF}}
  \end{aligned}
\end{equation}
while for the \gls{CCSD} model ($\mathcal{M} = 2$):
\begin{equation}\label{eq:source-ccsd}
  \begin{aligned}
  \s_{\mathrm{N}}(\tamp{}, \tbar{})_\mathrm{CCSD} &=
  \s_{\mathrm{N}}(\tamp{}, \tbar{})_\mathrm{CCS}
  + \braket{\tbar{1} | \BCHfirst{\s}{T_2} | \mathrm{HF}} \\
  &+ \braket{\tbar{2} | \BCHfirst{\s}{T_2} | \mathrm{HF}}
  + \braket{\tbar{2} | \commutator{\commutator{\s}{T_1}}{T_2} | \mathrm{HF}}
  \end{aligned}
\end{equation}
Eventually, within the \gls{CCSDT} model ($\mathcal{M} = 3$) four more
terms are added:
\begin{equation}\label{eq:source-ccsdt}
  \begin{aligned}
    \s_{\mathrm{N}}(\tamp{}, \tbar{})_\mathrm{CCSDT} &=
  \s_{\mathrm{N}}(\tamp{}, \tbar{})_\mathrm{CCSD}
  + \braket{\tbar{2} | \BCHfirst{\s}{T_3} | \mathrm{HF}} \\
  &+ \braket{\tbar{3} | \BCHfirst{\s}{T_3} | \mathrm{HF}} \\
  &+ \frac{1}{2}
  \braket{\tbar{3} | \commutator{\commutator{\s}{T_2}}{T_2} |
  \mathrm{HF}} \\
  &+ \braket{\tbar{3} | \commutator{\commutator{\s}{T_1}}{T_3} |
  \mathrm{HF}}
  \end{aligned}
\end{equation}

@@

\begarrowthought{Differentiation of the effective Lagrangian in Eq.
\eqref{eq:pted-cc}}
with respect to the variational parameters $\tamp{\mu_i}$,
$\tbar{\mu_i}$ and $\p$ yields the \acrshort{PTED}-\acrshort{CC}
equations:
\begin{subequations}\label{eq:pted}
  \begin{align}
   \tampEq{\mu_q}(\tamp{}, \tbar{}, \p)  &=
   \denom{\mu_q}\tamp{\mu_q} + \braket{\mu_q | \overline{\Phi} | \mathrm{HF}}
   + {}^t\p\braket{\mu_q | \overline{\s} | \mathrm{HF}}
             = 0 \label{eq:pted-cc-amplitudes}\\
   \tbarEq{\mu_q}(\tamp{}, \tbar{}, \p)
    &=
    \denom{\mu_q}\tbar{\mu_q} +
    \braket{\mathrm{HF} | \commutator{\overline{\Phi}}{\cluster{\mu_q}} | \mathrm{HF}} +
    \sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} |
    \commutator{\overline{\Phi}}{\cluster{\mu_q}} | \mathrm{HF}}
    \nonumber \\
    &+
    {}^t\p\braket{\mathrm{HF} | \commutator{\s}{\cluster{\mu_q}} | \mathrm{HF}}\delta_{\mu_q\mu_1}
    +
    {}^t\p\sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} |
    \commutator{\s}{\cluster{\mu_q}} | \mathrm{HF}} \nonumber \\
    &+
    {}^t\p\sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} | \commutator{\commutator{\s}{T}}{\cluster{\mu_q}} | \mathrm{HF}}
             = 0 \label{eq:pted-cc-multipliers}\\
    \Omega_\p(\tamp{}, \tbar{}, \p)
    &=
    \V\p + \s_{\mathrm{N}}(\tamp{}, \tbar{})_\mathcal{M} = 0
    \label{eq:pted-cc-polarization}
  \end{align}
\end{subequations}
A coupling of the amplitudes and multipliers equations is introduced by
the presence of the $\s_{\mathrm{N}}(\tamp{}, \tbar{})_\mathcal{M}$ expectation value
in the polarization equation.
This requires a proper macroiteration/microiteration self-consistency
scheme for its implementation.\autocite{Cammi2009-gu, Caricato2010-hx}
As a consequence, a single-point \acrshort{PTED}-\acrshort{CC} will suffer from a
$2\times$ prefactor in its computational cost with respect to an
\emph{in vacuo} \acrshort{CC} calculation.

\newthought{It is thus convenient to devise approximations} that are able to simplify the
Lagrangian and the governing equations by breaking or weakening the coupling.
While Eq. \eqref{eq:pted-cc-polarization} is \emph{directly} coupled to
the multipliers equation, since the source terms directly depends on
amplitudes and multipliers, it is only \emph{indirectly} coupled to the
amplitudes equation, where only the polarization enters.
Moreover, the leading terms in the correlated source operator
expectation value prominently involve the
singles cluster operator and the singles multiplier state.
Three approximations have currently been proposed: \gls{PTES}, \gls{PTE(S)}
and \gls{PTE}.

@@

  \hspace{-20pt}  {}^{\mathrm{PTES}}\lag(\tamp{}, \tbar{},
  \p)_\mathcal{M} &=
  {}^\mathrm{PTE}\lag(\tamp{}, \tbar{})_\mathcal{M} \\
  &+ \frac{1}{2}{}^t\p\V\p + {}^t\p\s_{\mathrm{N}}(\tamp{},
